On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-22 paper/paper3d.tex
re-writing intro

%\documentclass[12pt]{aastex}
\documentclass[iop, numberedappendix]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
%\usepackage{paralist}
\shorttitle{Non-ideal Gravitational instability}
\shortauthors{M-K.\ Lin \& K. Kratter}

@@

The standard metric for the (inverse) strength of disk self-gravity is
the \emph{Toomre parameter}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
\end{align}
\citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic frequency (which equals the rotation
frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
density and $G$ is the gravitational constant. The Toomre $Q$ is a
measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
the stabilzing effect of rotation ($\kappa$) and pressure 
($c_s$). This is evident from the dispersion relation, 
\begin{align}\label{classic_dr}
  s^2 =  2\pi G\Sigma|k| - \kappa^2 - c_s^2k^2,   
\end{align}
which relates the growth rate $s$ and radial wavenumber $k$ for local,
axisymmetric waves in a two-dimensional (2D, razor-thin), inviscid and
adiabatic (including isothermal) disk. Such disturbances are formally
unstable when the Toomre condition, $Q<1$, is satisfied.
Non-axisymmetric modes can develop for  larger but still order-unity
values of $Q$ . {\bf ref} 
%strictly speaking non axi *local* modes are stable -> only transient
%growth
%true non-axi instability needs global disk  model 
%non-axi GI at slightly larger Q, but still characterized by Q

@@

It is important to remember that the dispersion relation
Eq. \ref{classic_dr}  is derived from \emph{idealized} conditions: 
the disk is laminar and does not experience any net thermal
losses\footnote{Isothermal disks implicitly assume  
  heating and thermal losses are exactly balanced at all 
  times, so these do not cool in the current context.} 
(hereafter `cooling').  
However, real disks can cool, for example,
due to radiative losses. Accretion disks may also be turbulent, for
which its dynamic and thermodynamic effect is often modeled through an
effective viscosity \citep{shakura73,lin87,rafikov15}. {\bf 
  more refs here? non-sg refs} 
%in terms of AM transport and heating 
%assess GI in general

Despite this, the Toomre parameter, and its associated
dispersion relation, are still widely applied to realistic disk models
with viscosity and cooling \citep[e.g.][]{cossins09, kimura12}, 
even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
derived from inviscid disks without cooling.

However, as we emphasize in this work, the inclusion of non-ideal
physics, such as cooling and viscosity, modifies the dispersion
relation, Eq. \ref{classic_dr}, and hence the condition for GI. This 
means that the Toomre condition alone is insufficient to assess GI
in realistic disks. The goal of this work is to generalize the
analytic treatment of disk GI to allow a self-consistent discussion of
GI in realistic accretion disk models. Specifically, we consider the non-ideal
effects of viscosity and cooling on GI.

@@

\subsection{Cooling-driven gravitational instability}
Cooling reduces pressure (thermal) support against self-gravity. The cooling time 
$\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
to some pre-defined value $T_0$, which may be zero. That is, $\p_t T =
-(T-T_0)/\tcool$. It is often written as 
\begin{align} 
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time. This
type of cooling function, first applied by \cite{gammie01}, is a
simple parameterization that allows the full range of thermodynamic
response to be explored. 
{\bf physical origin of cooling in PPDs}

@@

We shall quantify how cooling removes pressure support to enable GI
even when $Q>1$.

We will find that cooling in fact changes the 
qualitative nature of GI, in that its mere presence can
formally lead to instability.

\subsection{Viscosity-driven gravitational instability}
%there is also viscous damping which stabilizes
Viscous disks can develop GI even when $Q>1$. This is because
viscosity compromises rotational support against self-gravity for
long-wavelength disturbances
\citep{lynden-bell74,willerding92,gammie96}. A similar effect  
occurs in dusty fluids where the required frictional forces are 
provided by dust-gas drag \citep{ward00, takahashi14}. In fact, this
is a mechanism to enhance particle clumping for planetesimational formation 
%frag of dust 
\citep{youdin11}. Then it is not unreasonable to expect an analogous
fragmentation of gaseous disks due to viscosity.

@@

>>>>

%\documentclass[12pt]{aastex}
\documentclass[iop, numberedappendix]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
%\usepackage{paralist}
\shorttitle{Non-ideal Gravitational instability}
\shortauthors{M.-K.\ Lin \& K. Kratter}

@@

The standard metric for the (inverse) strength of disk self-gravity is
the \emph{Toomre parameter}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
\end{align}
\citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic frequency (which equals the rotation
frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
density and $G$ is the gravitational constant. The Toomre $Q$ is a
measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
the stabilzing effect of rotation ($\kappa$) and pressure 
($c_s$). This is evident from the dispersion relation, 
\begin{align}\label{classic_dr}
  s^2 =  2\pi G\Sigma|k| - \kappa^2 - c_s^2k^2,   
\end{align}
which relates the growth rate $s$ and radial wavenumber $k$ for local,
axisymmetric waves in a two-dimensional (2D, razor-thin), inviscid and
adiabatic (including isothermal) disk. Such disturbances are formally
unstable when the Toomre condition, $Q<1$, is satisfied.
Non-axisymmetric modes can develop for larger, but still order-unity
values of $Q$ . {\bf ref} 
%strictly speaking non axi *local* modes are stable -> only transient
%growth
%true non-axi instability needs global disk  model 
%non-axi GI at slightly larger Q, but still characterized by Q

@@

It is important to remember that the popular dispersion relation, 
Eq. \ref{classic_dr},  is derived from \emph{idealized} conditions: 
the disk is laminar and does not experience any net thermal
losses\footnote{Isothermal disks implicitly assume  
  heating and thermal losses are exactly balanced at all 
  times, so these do not cool in the current context.} 
(hereafter `cooling').  
However, real disks can cool, for example,
due to radiative losses. Accretion disks may also be turbulent, for
which its dynamic and thermodynamic effect is often modeled through an
effective viscosity \citep{shakura73,lin87,rafikov15}. {\bf 
  more refs here? non-sg refs} 
%in terms of AM transport and heating 
%assess GI in general

Despite this, the Toomre parameter, and its associated
dispersion relation, are still widely applied to accretion disk models
with viscosity and cooling \citep[e.g.][]{cossins09, kimura12}, 
even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
derived from inviscid disks without cooling.

However, as we emphasize in this work, the inclusion of non-ideal
physics, such as cooling and viscosity, modifies the dispersion
relation, Eq. \ref{classic_dr}, and hence the condition for GI. This 
means that the Toomre condition alone is insufficient to assess GI
in realistic disks. The goal of this work is to generalize the
analytic treatment of disk GI to allow a self-consistent discussion of
GI in realistic accretion disk models. Specifically, we will examine how the
effects of cooling and viscosity can lead to GI by removing pressure
and rotational support, respectively, even when the Toomre condition
for instability is not satisfied.

@@

\subsection{Cooling-driven gravitational instability}
Cooling reduces pressure (thermal) support against self-gravity. The cooling time 
$\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
to some floor value $T_0$, which may be zero. That is, $\p_t T =
-(T-T_0)/\tcool$. It is often written as 
\begin{align} 
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time. This
type of cooling function, first applied by \cite{gammie01}, is a
simple parameterization that allows the full range of thermodynamic
response to be explored. 
{\bf physical origin of cooling in PPDs}

@@

We will quantify how cooling removes pressure support to enable GI
even when $Q>1$. For example, if perturbations are
cooled towards zero tempertaure, we find the above dispersion relation is
modified to read 
\begin{align}\label{cool_dr}
s^2 = 2\pi G\Sigma |k| - \kappa^2 - \gamma \left(\frac{\tcool s}{1 +
  \tcool s}\right)c_s^2k^2,
\end{align}
where $\gamma$ is the adiabatic index. This equation shows that
cooling increases the growth rate by reducing the magnitude of the
pressure term. In fact, Eq. \ref{cool_dr} implies that instability is
possible for any finite $Q$. This shows that cooling can change the
qualitative nature of GI compared to the classic Toomre condition.

%Eq. \ref{cool_dr} is a cubic in $s$ and differs qualitatively from the
%classic result, Eq. \ref{classic_dr}, which is quadratic in $s$. In
%fact, 
%We will find that cooling in fact changes the 
%qualitative nature of GI, in that its mere presence can
%formally lead to instability.

\subsection{Viscosity-driven gravitational instability}
%there is also viscous damping which stabilizes
Viscous disks can develop GI even when $Q>1$. This is because, as demonstrated below, viscosity compromises rotational support against
self-gravity for long-wavelength disturbances
\citep{lynden-bell74,willerding92,gammie96}. A similar effect   
occurs in dusty fluids where the required frictional forces are 
provided by dust-gas drag \citep{ward00, takahashi14}. In fact, this
is a mechanism to enhance particle clumping for planetesimational formation 
%frag of dust 
\citep{youdin11}. Then it is not unreasonable to expect an analogous
fragmentation of gaseous disks due to viscosity.

@@

\begin{align}
s \simeq \frac{\nu k^2\left(2\pi G\Sigma |k| -
  c_s^2k^2\right)}{\kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|}
\end{align}
The condition for instability,  
%\begin{align}
$  2\pi G \Sigma > c_s^2|k|$, is identical that obtained from 
Eq. \ref{classic_dr} with 
%\end{align}
